---
title: "Introduction to FloralCircadianR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to FloralCircadianR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6, fig.height = 4.5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
library(FloralCircadianR)
library(ggplot2)
library(dplyr)
library(shiny)
library(BIEN)
```
## Introduction

FloralCircadianR is an R package for analyzing and visualizing the circadian rhythms of flowering plants. The package allows users to:

Fetch plant species data for specific countries using the BIEN database.
Analyze and visualize the predominant circadian patterns (diurnal or nocturnal) for each country.
Create world maps to visualize species counts and circadian patterns.
This vignette demonstrates the workflow of FloralCircadianR and includes installation steps, core functions, and a practical example.

## Installation

To install the latest version of FloralCircadianR, use the following commands:

```{r}
install.packages("devtools")
library("devtools")
devtools::install_github("AUR0RACY/FloralCircadianR", build_vignettes = TRUE)
library(FloralCircadianR)
```

## Applications

The main functions of the FloralCircadianR package include:

1. Reading input data.
2. Fetching and analyzing species data.
3. Visualizing the results.
4. Interact with data through a user-friendly Shiny app.

Step 1: Read Input Data

First, prepare two CSV files:
  
- Circadian Data: Contains species names and their circadian properties (Diurnal or Nocturnal).
- Countries List: Contains a list of countries you are interested in.
    
The following functions read these files:
```{r}
# Example circadian data
circadian_data <- data.frame(
  scrubbed_species_binomial = c("Rosa", "Mangifera indica", "Betula papyrifera", "Eucalyptus globulus"),
  circadian_clock = c("Diurnal", "Nocturnal", "Diurnal", "Nocturnal")
)
# Example country data
countries <- c("Canada", "Brazil")
print(countries)
```

Step 2: Fetch Species Data

Use the fetch_species_for_countries() function to retrieve plant species data for the selected countries from the BIEN database.
```{r}
# Fetch species data for the selected countries
# Simulated fetch_species_for_countries example
species_data <- data.frame(
  country = rep(c("Canada", "Brazil"), each = 2),
  scrubbed_species_binomial = c("Rosa", "Mangifera indica", "Betula papyrifera", "Eucalyptus globulus")
)
print(species_data)
```

Step 3: Analyze Circadian Patterns

Use the find_country_circadian_pattern() function to determine the predominant circadian patterns for each country.

```{r}
# Simulated find_country_circadian_pattern example
country_circadian <-find_country_circadian_pattern(species_data, circadian_data)
print(country_circadian)
```

Step 4: Visualize the Results

Plot Species Count by Country
The plot_species_count_map() function creates a map showing the number of species found in each country. Countries are colored on a gradient from blue (fewer species) to red (more species).

```{r}
species_count_plot <- plot_species_count_map(species_data)
print(species_count_plot)
```
Plot Predominant Circadian Patterns
The plot_species_circadian_map() function creates a map showing the predominant circadian patterns in each country. Countries are colored yellow for "Diurnal" and blue for "Nocturnal."

```{r}
circadian_map <- plot_species_circadian_map(country_circadian)
print(circadian_map)
```

Step 5: Using the Shiny App
Launch the Shiny app to interact with the package:

```{r}
FloralCircadianR::runFloralCircadianR()
```
In the app:

1. Upload a CSV file containing the circadian data (scrubbed_species_binomial, circadian_clock).
2. Enter a list of countries of interest.
3. View and download visualizations for species count and circadian patterns.

The Shiny app provides a user-friendly interface for exploring circadian data.

## References

1. anjalisilva. (2019). GitHub - anjalisilva/TestingPackage: R Package Illustrating Components of an R package for BCB410H - Applied Bioinformatics (2019-2023), University of Toronto, Canada. GitHub. https://github.com/anjalisilva/TestingPackage

2. OpenAI. (2024). ChatGPT. ChatGPT; OpenAI. https://chatgpt.com/

3. R Core Team (2023). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. https://www.R-project.org/

4. RBIEN – Botanical Information and Ecology Network. (2024). Ucsb.edu. https://bien.nceas.ucsb.edu/bien/tools/rbien/

5. Wickham, H. (2019). Create Elegant Data Visualisations Using the Grammar of Graphics. Tidyverse.org. https://ggplot2.tidyverse.org/

6. Wickham, H., François, R., Henry, L., Müller, K., & Vaughan, D. (2019). A Grammar of Data Manipulation. Tidyverse.org. https://dplyr.tidyverse.org/

7. Chang, W., Cheng, J., Allaire, J., Xie, Y., & McPherson, J. (2024). Shiny: Web Application Framework for R. R package version 1.9.1. https://CRAN.R-project.org/package=shiny